Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to US Midwest agroecosystems

### 2021-08-23 Manuscript/Manuscript.tex
as

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is currently
considered one of the most economically damaged weed species to cropping
systems in the United States. Unmanaged Palmer amaranth in competition
for water, light and nutrients can drastically impact on crop yields
(Berger et al., 2015). For example, Palmer amaranth is documented with
potential to reduce 91\%, 68\%, and 54\% of corn (Massinga et al.,
2001), soybean (Klingaman and Oliver, 1994), and cotton (Morgan et al.,
2001) yields, respectively. Moreover, Palmer amaranth has showed a
remarkable capacity to evolve resistance to herbicides. To date, Palmer
amaranth has evolved resistance to eight herbicide sites of action
(Heap, 2021), increasing the weed management complexity (Lindsay et al.,
2017). Thus, Palmer amaranth poses an economical risk to sustainable
agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with species
biology. Farm mechanization, conservation agriculture (e.g., no-till),
and reliance on herbicides for weed management are the main
human-mediated selection of Palmer amaranth into cropping systems (Ward
et al., 2013). On the other hand, Palmer amaranth is a prolific seed
producer with a C4 photosynthetic apparatus (Wang et al., 1992). With a
dioecy nature, Palmer amaranth male and female plants are obligate
outcrosser species, increasing the chances of exchanging adaptive traits
among plants (Oliveira et al., 2018). Also, Palmer amaranth small seed
size (e.g, 1 mm) tend to thrive in no-tillage systems (Price et al.,
2011), and spread across locations through farm equipment (Sauer, 1972),
manure (Hartzler and Anderson, 2016), animals (Farmer et al., 2017), and
plant propagules (Yu et al., 2021). The dispersal capacity of Palmer
amaranth make the species one of the most successful cases of weed
adaption to cropping systems.

Light and temperature are likely the main environment requirements for
Palmer amaranth successful grow and development. Palmer amaranth is
reported with an extended germination period (Jha et al., 2010).
Germination of Palmer amaranth is triggered by 18 C soil temperature
(Keeley et al., 1987), and optimal germination and biomass production
occur at 35/30 C day and night temperatures (Guo and Al-Khatib, 2003).
Palmer amaranth germination is also human-mediated by tillage timings
and preemergence-applied herbicides (Chahal et al., 2021). Water has not
shown to limit Palmer amaranth fitness. Under continuous water stress,
Palmer amaranth survived and produced at least 14000 seeds plant-1
(Chahal et al., 2018). Seeds from Palmer amaranth growing with limited
water conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). Growing conditions and management practices
also influence Palmer amaranth sex dimorphism and flowering pattern
(Korres et al., 2017; Rumpa et al., 2019). The continuous global
temperature warming can impact agriculture and promote niches for Palmer
amaranth invasion/adaptation into new environments. Currently, it is
estimated that the greatest climatic risk of Palmer amaranth
establishment are agronomic crops in Australia and Sub-Sahara Africa
(Kistner and Hatfield, 2018). Temperature is a key factor limiting
Palmer amaranth expansion to cooler geographies (Briscoe Runquist et
al., 2019); however, under future climate change Palmer amaranth is
likely to expand northward into Canada and Northern Europe (Kistner and
Hatfield, 2018; Briscoe Runquist et al., 2019).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the US, Palmer amaranth is well established at cotton
fields (Garetson et al., 2019) and non-crop areas (Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021), Michigan (Kohrt et al.,
2017), Connecticut (Aulakh et al., 2021), and potential damaged to
soybean fields in Illinois is reported (Davis et al., 2015). In Iowa, a
study showed that Palmer amaranth is still not well adapted compared to
waterhemp (\emph{Amaranthus tuberculatus}) (Baker, 2021). Successful
cases of Palmer amaranth invasion and near to eradication is documented
in Minnesota (Yu et al., 2021). No Palmer amaranth actively growing was
found in Canada; however, Palmer amaranth seeds was detected in sweet
potato slips (Page et al., 2021). Nonetheless, it seems fated the need
to manage Palmer amaranth in agronomic crops throughout multiple
environments in the near future. Strategies on Palmer amaranth
management should encompass the agroecosystem level but not only
attempts to eradicate the weed. Most tactics to manage Palmer amaranth
are based on technology fixes (Scott, 2011), which are short-term (e.g.,
herbicide and/or tillage) rather than long-term weed management. Palmer
amaranth management should be built on minimizing the species ability to
adapt, grow and develop into agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to
have higher impact on cotton yields than late established plants (MacRae
et al., 2013). In the northern states, Palmer amaranth impact on the
agroecossystem is recent. Studies investigating Palmer amaranth in those
locations is limited due to the plant classification as noxious weed
species (Yu et al., 2021). Nonetheless, the continuous Palmer amaranth
dispersal and potential establishment across northern United States is
concerning and warrant investigations on species morphology in such
environments. Understanding Palmer amaranth biology and growing
strategies under different agroecossystems can enhance our knowledge on
species adaptation. It can also aid on designing proactive and
ecological tactics to limit the species range expansion, reduce its
negative impact, and design resilient and sustainable farming systems
(MacLaren et al., 2020). Therefore, the objective of this study was to
investigate the flowering pattern, gender, biomass production, and
height of Palmer amaranth growing under corn, soybean and bareground at
two timings across five locations in the United States Midwest.

@@

The study was performed with a Palmer amaranth accession (Per1) from
Perkins County, Nebraska. Per1 accession collection is documented with
no reported herbicide resistance (Oliveira et al., 2021). Three weeks
prior to the field experiment, seeds were planted in plastic trays
containing potting-mix. Emerged seedlings (1 cm) were transplanted into
200 cm-3 plastic pots (a plant pot-1). Palmer amaranth seedlings were
supplied with adequate water and kept under greenhouse conditions at
Arlington, Clay Center, Lincoln, and Macomb; and kept outdoors in Grant.
Palmer amaranth seedlings were kept under greenhouse/outdoors until the
onset of the experiment (2-3 leaf stage/5 to 8 cm height).

@@

The experiment was conducted in 2018 and 2019 under field conditions at
five locations: Arlington, WI (43°18'N, 89°29'W), Clay Center, NE ('N,
'W), Grant (('N, 'W)), Lincoln (('N, 'W)), and Macomb (('N, 'W)).

Fields were conventionally tilled and disked prior to planting. Corn
hybrid and soybean varieties were planted in 76-cm row spacing (Table
1). Monthly mean air temperature and sum precipitation were obtained
using Daymet weather data from June through September across the five
locations in 2018 and 2019 (Correndo et al., 2021) (Figure 1)

@@

\multirow{-3}{*}{\raggedright\arraybackslash } & Weed control & glyphosate / S-metolachor & S-metolachlor + trazine + mesotrione, + bicyclopyrone &  &  & \\

@@

\multirow{-3}{*}{\raggedright\arraybackslash } & Weed control & glyphosate / S-metolachor & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

@@

The field experimental unit were six adjacent 9.1 m wide (12 rows at
72.2 cm row spacing) by 10.7 m long. Each experimental unit was planted
with corn or soybean, or under bareground condition. Palmer amaranth
seedlings (potting mix + two seedlings) were and gently transferring to
the ground (6 cm deep and 8 cm wide). Twenty-four plants were
equidistantly placed (0.76 m apart) between rows within each
agroecossystems. After a week, one was eliminated and one was kept.
There were two transplant timing: first (June 1) and second (July 1).
There were 24 Palmer amaranth plants in each experimental unit, with a
total of 144 plants for each location. The study was repeated twice.

After transplanting, Palmer amaranth flowering was monitored until the
end of the study. When a plant started flowering, the day was recorded,
plant sex was identified as male or female, and plant height was
measured from soil surface to the plant top. Then, aboveground plant
biomass was harvest near soil surface and oven dried at 65 C until
reaching constant weight before the weight of biomass (g plant 1) was
recorded.

@@

The statistical analyses were performed using R statistical software
version 4.0.1. Data across locations and year were combined.

The cumulative Palmer amaranth flowering estimation was determined using
a asymmetrical three parameter log logistic Weibull model of the drc
package (Ritz et al., 2015).

\(Y(x) = 0 + (d-0) exp (-exp(b(log(x)-e)))\)

@@

The doy for 10, 50, and 90\% Palmer amaranth cumulative flowering were
determined using the \emph{ED} function of drc package. Also, the 10,
50, and 90\% Palmer amaranth cumulative flowering were compared among
agroecossystems and timings using the \emph{EDcomp} function of drc
package. The EDcomp function compares the ratio of cumulative flowering
using t-statistics, where P-value \textless{} 0.05 indicates that we
fail to reject the null hypothesis.

Palmer amaranth gender was fitted to a binary logistic regression
(Bangdiwala, 2018). Binary logistic regression is used for predicting
binary classes, such as the probability of female in dioecious plants.
Prior to the analysis, all missing values were removed from the dataset.
Also, data from Grant was not used in this analysis due to the uniform
plant harvesting at that location. The whole dataset was splitted into
80\% train and 20\% test data. The 80\% train is used for the model
training and the rest 20\% is used for checking how the model
generalized on unseen dataset. With 80\% dataset, a binary response
variable, male (0) and female (1), was fitted to a generalized linear
model (\emph{glm} function) including day of year harvest, height,
weight, crop and month as independent variables. The model family was
binomial with a logit function. The model fit was assessed through
pseudo R squared values (McFadden, Cox and Snell, Cragg and Uhler) and
likelihood ratio using \emph{nagelkerke} function (``rcompanion''
package). The marginal effects computation was performed with Average
Marginal Effects at every observed value of x and average across the
results (AMEs) (Leeper, 2017) using \emph{margins} function from
``margins'' package. The rest 20\% dataset was predicted using
\emph{predict} function with a cutoff estimation for male or female
using \emph{performance} function. The model quality prediction from the
classification algorithm was measure with precision (\emph{precision}
function), recall (\emph{recall} function) and F1 score (\emph{f\_meas}
function) using ``yardstick'' package. The precision determines the
accuracy of positive predictions (female plants), recall determines the
fraction of positives that were correctly identified, and F1 score is a
weighted harmonic mean of precision and recall with the best score of 1
and the worst score of 0. F1 score conveys the balance between the
precision and the recall.

Palmer amaranth height and biomass were performed with a linear mixed
model using \emph{lmer} function from ``lme4'' package (Bates et al.,
2015). Plant height and biomass were transformed to meet model
assumption of normality. In the model, agroecosystem (crop, soybean,
bareground) was the fixed effect and year nested with location the
random effects. Analysis of variance was performed with \emph{anova}
function from ``car'' package (Fox and Weisberg, 2018). Marginal means
and compact letter display were estimated with \emph{emmeans} and
\emph{cld} from packages ``emmeans'' and ``multcomp'' (Hothorn et al.,
2008).

@@

Palmer amaranth accumulated more biomass when growing in bareground
compared to Palmer amaranth growing in soybean and corn (figure 2A). At
first cohort time, Palmer amaranth biomass was 75.5 g plant-1, 28.3 g
plant-1 and 16.3 g plant-1 in bareground, soybean and corn,
respectively. At second cohort timing, Palmer amaranth produced 62.6 g
plant in bareground, followed by 6.3 g plant in soybean, and 1.4 g plant
in corn.

Palmer amaranth height was more uniform across cohort timings, except
when growing in corn (figure 2B). Palmer amaranth achieved 69.2 cm tall
when growing at first cohort timing in bareground, which was not
different to 70.7 cm tall at second cohort timing (P = 0.74). In
addition, no difference in Palmer amaranth height (69.3 cm) was found in
soybean at first cohort compared to bareground (P \textgreater{} 0.75).
At second cohort in soybean, Palmer amaranth was near to 10 cm lower
compared to its first cohort time (P = 0.04). The tallest and smallest
Palmer amaranth plants were found in corn. Palmer amaranth reached 85.2
cm tall at first cohort and 38.2 cm at second cohort timings.

@@

Palmer amaranth growing in corn resulted in a longer flowering pattern
compared to bareground and soybean at first cohort (figure3A).
Nonetheless, the 10\% cumulative Palmer amaranth flowering in soybean,
bareground and corn occurred at the end of June. Palmer amaranth reached
10\% flowering in soybean at doy 180, which was slightly different from
bareground (doy 180.9; \emph{P} = 0.01) and corn (doy 181.7; \emph{P} =
0.00). The 50\% Palmer amaranth cumulative flowering occurred in July.
For example, Palmer amaranth reached 50\% flowering in bareground at doy
193.4, followed by soybean (doy 194.8), corn (doy 206.6). Similar trend
was observed at 90\% Palmer amaranth cumulative flowering. Palmer
amaranth growing in corn reached 90\% flowering at doy 252.6 (early
September), which was 37.8 and 32.2 days after Palmer amaranth 90\%
flowering in bareground and soybean, respectively.

Palmer amaranth cumulative flowering at second cohort ranged from mid
July to mid September (figure3B). Palmer amaranth growing in bareground
resulted in earlier flowering time compared to soybean and corn. Palmer
amaranth growing in bareground reached 10\%, 50\%, and 90\% flowering
time at day 203.8, 214.4, and 232.2, respectively. Palmer amaranth
growing in soybean reached 10\% flowering at doy 210.9, which was 6 days
prior to corn (\emph{P}-value = 0.00). Similar trend was observed at
50\% flowering, whereas Palmer amaranth reached 50\% flowering in corn
(doy 233.0) 4 days after soybeans (doy 228.9; \emph{P} = 0.00). The 90\%
Palmer amaranth cumulative flowering occurred at same day in corn
(260.9) and soybean (260.5; \emph{P} = 0.66).

@@

The model fit was 0.23, 0.32, 0.40 with using pseudo R squared test from
McFadden, Cox and Snell, and Cragg and Uhler, respectively. The
likelihood ratio test showed a p-value of \textless{} 0.00. The average
marginal effects showed that Palmer amaranth growing in corn resulted in
14.5\% less females plants (Table 2). Moreover, increasing a unit doy
increases the probability of having a female plant by 0.4\% (Table 2 and
Figure4A). Similar trend is observed to weight as well as height,
whereas the probability of being female increase by 0.2\% (Figure 4B)
and 0.1\% (Figure 4C) when a unit of weight (g) and height (cm)
increases, respectively.

@@

The model accuracy evaluation accuracy in the unseen 20\% dataset was
0.62 with a cutoff value for female and male plants of 0.43. The model
classification showed a precision of 0.64, recall of 0.66, and a F means
score of 0.65. In addition, the area under the curve was 0.64.

@@

Our study showed that Palmer amaranth biomass, height, flowering pattern
and gender varied within agroecossystems and cohort timings. In general,
Palmer amaranth produced more biomass and taller plants when growing at
first cohort rather than second cohort. At first cohort, resources
(e.g., soil nutrients) and conditions (e.g., light) were more timely
available for both species, crop and weed. High biomass and taller
Palmer amaranth plants are likely a weed strategy to compete for light
in between crop rows in absence of canopy. In such condition, Palmer
amaranth showed an extraordinary plasticity to adapt upon the
agroecosystem. This is evident when comparing Palmer amaranth canopy
shape, and its extended flowering pattern when growing into corn
compared to soybean. The Palmer amaranth competition strategy was to
mimic the crop grow and development (Figure 4). These results suggests
that Palmer amaranth can fast evolve life-history traits to adapt into
agroecossystems and cultural practices, including nitrogen fertilization
(Bravo et al., 2018). Our results highlight the Palmer amaranth as a
threat to agroecossystems as breeding more competitive crop varieties is
likely to select more competitive biotypes (Bravo et al., 2017).

Palmer amaranth grow and development at second cohort was limited due to
the crop competitive ability at advanced development stages. Palmer
amaranth was transplanted when corn canopy was nearly closed, which
reduced Palmer amaranth competitiveness. As a result, Palmer amaranth
height and biomass was lower compared to its first cohort. Under crop
canopy (e.g., second cohort), Palmer amaranth flowering pattern was near
to similar in corn and soybean. Palmer amaranth growing without crop
competition produced highest amounts of biomass and less extended
flowering pattern. The Palmer amaranth strategy in bareground was to
invest biomass in growing plant width and height. Nonetheless, Palmer
amaranth produced 21\% less biomass in second cohort compared to first
cohort timing. In a bareground study, early emerged Palmer amaranth
without competition was 50\% taller than late emerged plants (Webster
and Grey, 2015). These results suggests that crop competition is not the
only factor limiting late Palmer amaranth establishment. The limited
growth of Palmer amaranth at second cohort is likely a reduced plant
response to thermal units (e.g, growing degree days). It is hypothesize
that reduced day length contributed to smaller plants at second cohort
as well as shorter flowering period. A study in North Carolina and
Illinois predicted that less than 10\% Palmer amaranth seedlings
emergence occurred after June (Piskackova et al., 2021). In addition,
Palmer amaranth negative impact on soybean (Korres et al., 2020) and
cotton (Webster and Grey, 2015) yields was higher when plants were
established near to crop planting.

>>>>

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is currently
ranked as one of the most economically detrimental weed species to
cropping systems in the United States (ADD CITATION; WSSA Survey
Perhaps). Unmanaged Palmer amaranth plants compete for water, light, and
nutrients, which can drastically impact crop yields (Berger et al.,
2015). For example, Palmer amaranth has been documented to reduce up to
91\%, 68\%, and 54\% corn (Massinga et al., 2001), soybean (Klingaman
and Oliver, 1994), and cotton (Morgan et al., 2001) yields,
respectively. Moreover, Palmer amaranth has shown a remarkable capacity
to evolve resistance to herbicides. To date, Palmer amaranth has evolved
resistance to eight herbicide sites of action (Heap, 2021), increasing
the weed management complexity (Lindsay et al., 2017). Thus, Palmer
amaranth poses an economical and environmental risk to sustainable
agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to the
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with plant
biology. Farm mechanization, adoption of conservation agriculture (e.g.,
no-till), and reliance on herbicides for weed management are the main
human-mediated selection of Palmer amaranth into cropping systems (Ward
et al., 2013). On the other hand, Palmer amaranth is a prolific seed
producer with a C4 photosynthetic apparatus (Wang et al., 1992). With a
dioecy nature, Palmer amaranth male and female plants are obligate
outcrosser species, increasing the chances of exchanging adaptive traits
among plants (Oliveira et al., 2018). Also, Palmer amaranth small seed
size (e.g, 1 mm) tend to thrive in no-tillage systems (Price et al.,
2011), and spread across locations through farm equipment (Sauer, 1972),
manure (Hartzler and Anderson, 2016), animals (Farmer et al., 2017). The
dispersal capacity of Palmer amaranth makes the species one of the most
successful cases of weed adaptation to current cropping systems.

Light and temperature are likely the main environment requirements for
Palmer amaranth successful grow and development (CITATION). Palmer
amaranth is reported with an extended germination period (Jha et al.,
2010). Germination of Palmer amaranth is triggered by 18 C soil
temperature (Keeley et al., 1987), and optimal germination and biomass
production occur at 35/30 C day and night temperatures (Guo and
Al-Khatib, 2003). In addition, Palmer amaranth establishment is
human-mediated by tillage timings and preemergence-applied herbicides
(Chahal et al., 2021), which can result in weed germination shifts
(Sbatella and Wilson, 2010). Water has not shown to limit Palmer
amaranth fitness. Under continuous water stress, Palmer amaranth
survived and produced at least 14000 seeds plant-1 (Chahal et al.,
2018). Seeds from Palmer amaranth growing with limited water conditions
were heavier, less dormant, and prompt for germination (Matzrafi et al.,
2021). Growing conditions and management practices also influence Palmer
amaranth sex dimorphism and flowering pattern (Korres et al., 2017;
Rumpa et al., 2019). Therefore, Palmer amaranth has shown plasticity to
evolve and adapt under selection pressure. The continuous global
temperature warming is likely impact agriculture and promote niches for
Palmer amaranth invasion/adaptation into new habitats. Currently, it is
estimated that the greatest climatic risk of Palmer amaranth
establishment are agronomic crops in Australia and Sub-Sahara Africa
(Kistner and Hatfield, 2018). Temperature is a key factor limiting
Palmer amaranth expansion to cooler geographies (Briscoe Runquist et
al., 2019); however, under future climate change Palmer amaranth is
likely to expand northward into Canada and Northern Europe (Kistner and
Hatfield, 2018; Briscoe Runquist et al., 2019).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the United States, Palmer amaranth is well established
in the Cotton Belt (Garetson et al., 2019; Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021). There are some reported
cases of Palmer amaranth in Michigan (Kohrt et al., 2017) and
Connecticut (Aulakh et al., 2021). Also, it is estimated that Palmer
amaranth can cause high damage to soybean fields in Illinois (Davis et
al., 2015), which is concerning as soybean along with corn make most of
US Midwest agronomic hectares. In Iowa, a study showed that Palmer
amaranth is still not well adapted compared to waterhemp
(\emph{Amaranthus tuberculatus}) (Baker, 2021). Invasion and successful
eradication of Palmer amaranth is documented in Minnesota (Yu et al.,
2021). Palmer amaranth infestations have not been detected in Canada;
however, Palmer amaranth seeds were detected in sweet potato slips in
the country (Page et al., 2021). Palmer amaranth is still not as well
adapted and established to Northern as it is in the Southern North
America. Therefore, its range of expansion into new habitats is
increasing. It seems fated the need to manage Palmer amaranth in
agronomic crops throughout multiple environments in the near future.
Strategies on Palmer amaranth management should encompass the
agroecosystem level but not only attempts to eradicate the weed. Most
tactics to manage Palmer amaranth are based on technology fixes (Scott,
2011), which are short-term (e.g., herbicide and/or tillage) rather than
long-term weed management. Palmer amaranth management should be built on
minimizing the species ability to adapt, grow and develop into
agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to
have a higher impact on cotton yields compared to late established
plants (MacRae et al., 2013). In the northern states, Palmer amaranth
impact on the agroecosystem is recent. Studies investigating Palmer
amaranth in those locations are limited due to the plant classification
as noxious weed species in some northern states (Yu et al., 2021).
Nonetheless, the continuous Palmer amaranth dispersal and potential
establishment across the northern United States is concerning and
warrants investigations on species morphology in such environments.
Understanding Palmer amaranth biology and growing strategies under
different agroecosystems can enhance our knowledge on species adaptation
and management practices. It can also aid in designing proactive and
ecological tactics to limit the species range expansion, reduce its
negative impact, and developing resilient and sustainable farming
systems (MacLaren et al., 2020). Therefore, the objective of this study
was to investigate the flowering pattern, gender, biomass production,
and height of Palmer amaranth cohorts growing under corn, soybean and
bareground across five locations in the United States Midwest.

@@

A Palmer amaranth accession (Per1) from Perkins County, Nebraska. with
no reported herbicide resistance according to Oliveira et al. (2021) was
selected for this study. Three weeks prior to the onset of the field
experiments, seeds were planted in plastic trays containing potting-mix.
Emerged seedlings (1 cm) were transplanted into 200 cm-3 plastic pots (a
plant pot-1). Palmer amaranth seedlings were supplied with adequate
water and kept under greenhouse conditions at Arlington, Clay Center,
Lincoln, and Macomb; and kept outdoors in Grant until the 2-3 leaf stage
(5 to 8 cm height) when they were transported to the field.

@@

The experiment was conducted in 2018 and 2019 under field conditions at
five locations: Arlington, WI (43°18'N, 89°29'W), Clay Center, NE ('N,
'W), Grant, NE (('N, 'W)), Lincoln, NE (('N, 'W)), and Macomb, IL (('N,
'W)).

Fields were conventionally tilled prior to crop planting. Corn hybrid
and soybean varieties were planted in 76-cm row spacing (Table 1).
Monthly mean air temperature and total precipitation were obtained using
Daymet weather data from June through September across the five
locations in 2018 and 2019 (Correndo et al., 2021) (Figure 1)

@@

 & Weed control & glyphosate / S-metolachor & S-metolachlor + trazine + mesotrione, + bicyclopyrone &  &  & \\

 & Stage 1 cohort & V2-3 &  &  &  & \\

\multirow{-5}{*}{\raggedright\arraybackslash } & Stage 2 cohort & V6-7 &  &  &  & \\

@@

 & Stage 1 cohort & V1-2 &  &  &  & \\

 & Stage 2 cohort & V5-6 &  &  &  & \\

\multirow{-5}{*}{\raggedright\arraybackslash } & Weed control & glyphosate / S-metolachor & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

@@

The field experimental units were three adjacent 9.1 m wide (12 rows at
76.2 cm row spacing) by 10.7 m long. The experimental design were
arranged in factorial design with three crops, two transplanting times
simulating two cohorts, repeated across five locations. Each
experimental unit was planted with corn, soybean, or kept under
bareground. The two transplanting timings were June 1 (first cohort) and
July 1 (second cohort). Palmer amaranth seedlings (potting mix + two
seedlings) were transplanted (6 cm deep and 8 cm wide). Forty-eight
plants were equidistantly placed (0.76 m apart) between rows within each
crop (Figure 2). After a week, one plant was eliminated and one was
kept, resulting in 24 plants per experimental unit and transplanting
time (Figure 2). When needed, Palmer amaranth plants were supplied with
water during the first week after transplanting.

After transplanting, Palmer amaranth flowering was monitored until the
end of the study. When a plant flowered, the day was recorded, plant sex
was identified (male or female), plant height was measured from soil
surface to the top of plant. Also, aboveground plant biomass was harvest
near soil surface and oven dried at 65 C until reaching constant weight
before weighing (g plant-1 was recorded).

Plants had to be harvested at flowering because Palmer amaranth is
neither endemic in Wisconsin nor in Illinois. In our study, all
locations follow the methodology of plant harvest at flowering
initiation, except in Grant, NE. In this location, all Palmer amaranth
plants were harvest at once on July 06, 2018 and 2019 (first cohort),
and on August 17, 2018 and on July 31, 2019 (second cohort).

@@

The statistical analyses were performed using R statistical software
version 4.0.1.

Analyses of Palmer amaranth height and biomass were performed with a
linear mixed model using \emph{lmer} function from ``lme4'' package
(Bates et al., 2015). Plant height and biomass were log transformed to
meet model assumption of normality. In the model, crop (bareground,
corn, soybean) was the fixed effect and year nested with location the
random effects. Analysis of variance was performed with \emph{anova}
function from ``car'' package (Fox and Weisberg, 2018). Marginal means
and compact letter display were estimated with \emph{emmeans} and
\emph{cld} from packages ``emmeans'' and ``multcomp'' (Hothorn et al.,
2008).

The cumulative Palmer amaranth flowering estimation was determined using
an asymmetrical three parameter log logistic Weibull model of the drc
package (Ritz et al., 2015).

\[Y(x) = 0 + (d-0) exp (-exp(b(log(x)-e)))\]

@@

The doy for 10, 50, and 90\% Palmer amaranth cumulative flowering were
determined using the \emph{ED} function of drc package. Also, the 10,
50, and 90\% Palmer amaranth cumulative flowering were compared among
crops and cohorts using the \emph{EDcomp} function of drc package. The
EDcomp function compares the ratio of cumulative flowering using
t-statistics, where P-value \textless{} 0.05 indicates that we fail to
reject the null hypothesis.

Palmer amaranth sex was fitted to a binary logistic regression
(Bangdiwala, 2018). Binary logistic regression is used for predicting
binary classes, such as the probability of a plant being female in a
dioecious species. Prior to the analysis, all missing values were
removed from the dataset. Also, data from Grant was not used in this
analysis due to the uniform plant harvesting at that location. The
complete dataset was splitted into 80\% train and 20\% test data. The
80\% train is used for the model training and the 20\% test is used for
checking how the model generalized on unseen dataset. With 80\% dataset,
a binary response variable, male (0) and female (1), was fitted to a
generalized linear model (\emph{glm} function) including day of year
harvest, height, weight, crop and month as independent variables. The
model family was binomial with a logit function. The model fit was
assessed through pseudo R-squared values (McFadden, Cox and Snell, Cragg
and Uhler) and likelihood ratio using \emph{nagelkerke} function
(``rcompanion'' package). The marginal effects computation was performed
with Average Marginal Effects at every observed value of x and average
across the results (AMEs) (Leeper, 2017) using \emph{margins} function
from ``margins'' package. The rest 20\% dataset was predicted using
\emph{predict} function with a cutoff estimation for male or female
using \emph{performance} function. The model quality prediction from the
classification algorithm was measured with precision (\emph{precision}
function), recall (\emph{recall} function) and F1 score (\emph{f\_meas}
function) using the ``yardstick'' package. The precision determines the
accuracy of positive predictions (female plants), recall determines the
fraction of positives that were correctly identified, and F1 score is a
weighted harmonic mean of precision and recall with the best score of 1
and the worst score of 0. F1 score conveys the balance between the
precision and the recall.

@@

Palmer amaranth plants accumulated more biomass when growing in
bareground compared to plants growing in soybean and corn (figure 3A).
Palmer amaranth plants in the first cohort produced 75.5, 28.3 g, and
16.3 g plant-1 in bareground, soybean and corn, respectively. Plants
from the second cohort produced 62.6 g plant in bareground, followed by
6.3 g plant in soybean, and 1.4 g plant in corn.

Palmer amaranth height was more uniform across cohort timings, except
when growing in corn (figure 3B). Palmer amaranth plants from the first
cohort were on average 69.2 cm tall in bareground, which was not
different from the 70.7 cm tall plants from the second cohort timing (P
= 0.74). In addition, no difference in Palmer amaranth height (69.3 cm)
was detected from first cohort plants in soybean to first and second
cohort plants in bareground (P \textgreater{} 0.75). Palmer amaranth
plants from the second cohort were nearly 10 cm lower compared to the
first cohort in soybeans (P = 0.04). The tallest (first cohort) and
smallest (second cohort) Palmer amaranth plants were found in corn.
Palmer amaranth reached 85.2 and 38.2 cm tall, respectively.

@@

Palmer amaranth plants from the first cohort growing in corn resulted in
a longer flowering window compared to plants growing in bareground and
soybean (Figure 4A). The 10\% cumulative Palmer amaranth flowering in
soybean, bareground and corn occurred at the end of June. Palmer
amaranth reached 10\% flowering in soybean, bareground and corn at doy
180, 180.9 and 181.7, respectively. The 50\% Palmer amaranth cumulative
flowering occurred in July. Palmer amaranth reached 50\% flowering in
bareground, soybean and corn at doy 193.4, 194.8, and 206.6,
respectively. Similar trend was observed at 90\% Palmer amaranth
cumulative flowering. Palmer amaranth reached 90\% flowering at doy
252.6 in corn (early September), which was 38 and 32 days after reaching
90\% flowering in bareground and soybean, respectively.

Palmer amaranth cumulative flowering at the second cohort ranged from
mid July to mid September (Figure 4B). Palmer amaranth growing in the
bareground resulted in earlier flowering time compared to soybean and
corn. Palmer amaranth growing in bareground reached 10\%, 50\%, and 90\%
flowering time at day 203.8, 214.4, and 232.2, respectively. Palmer
amaranth growing in soybean reached 10\% flowering at doy 210.9, which
was 6 days prior to corn (\emph{P}-value = 0.00). Similar trend was
observed at 50\% flowering, whereas Palmer amaranth reached 50\%
flowering in corn (doy 233.0) 4 days after soybeans (doy 228.9; \emph{P}
= 0.00). The 90\% Palmer amaranth cumulative flowering occurred at same
day in corn (260.9) and soybean (260.5; \emph{P} = 0.66).

@@

The model fit was 0.23, 0.32, 0.40 using pseudo R-squared test from
McFadden, Cox and Snell, and Cragg and Uhler, respectively. The
likelihood ratio test showed a p-value of \textless{} 0.00. The average
marginal effects showed that Palmer amaranth growing in corn resulted in
14.8\% less females plants (Table 2). Moreover, increasing a unit doy
increases the probability of having a female plant by 0.4\% (Table 2 and
Figure4A). Similar trend is observed for weight as well as height,
whereas the probability of being female increase by 0.2\% (Figure 4B)
and 0.1\% (Figure 4C) when a unit of weight (g) and height (cm)
increases, respectively.

@@

The model accuracy evaluation accuracy in the 20\% test dataset was 0.62
with a cutoff value for female and male plants of 0.43. The model
classification showed a precision of 0.64, recall of 0.66, and a F means
score of 0.65. In addition, the area under the curve was 0.64.

@@

Our study showed that Palmer amaranth biomass, height, flowering pattern
and gender varied within agroecossystems and cohort timings. In general,
first cohort of Palmer amaranth plants were heavier and taller when
compared to the second cohort. At first cohort, resources (e.g., soil
nutrients) and conditions (e.g., light) were more timely available for
the species. High biomass and taller Palmer amaranth plants are likely a
weed strategy to compete for light in between crop rows in absence of
canopy. In such conditions, Palmer amaranth showed an extraordinary
plasticity to adapt upon the agroecosystem. This is evident when
comparing Palmer amaranth canopy shape, and its extended flowering
pattern when growing into corn compared to soybean. The Palmer amaranth
competition strategy was to mimic the crop grow and development (Figure
6). These results suggests that Palmer amaranth can fast evolve
life-history traits to adapt into cropping systems and cultural
practices, which was also showed in a study varying nitrogen
fertilization (Bravo et al., 2018). Our results highlight the Palmer
amaranth as a threat to field crops as breeding more competitive crop
varieties is likely to select more competitive biotypes (Bravo et al.,
2017).

Palmer amaranth grow and development in second cohort was limited due to
the crop competitive ability at advanced development stages. Palmer
amaranth was transplanted when corn canopy was nearly closed, which
reduced Palmer amaranth competitiveness. As a result, Palmer amaranth
height and biomass was lower compared to its first cohort. Under crop
canopy (e.g., second cohort), Palmer amaranth flowering window was near
to similar in corn and soybean. Palmer amaranth growing without crop
competition produced the highest amounts of biomass and less extended
flowering window. The Palmer amaranth strategy in bareground was to
invest biomass in growing plant width and height. Nonetheless, Palmer
amaranth produced 21\% less biomass in second cohort compared to first
cohort timing. In a bareground study, early emerged Palmer amaranth
without competition was 50\% taller than late emerged plants (Webster
and Grey, 2015). These results suggest that crop competition is not the
only factor limiting late Palmer amaranth establishment. The limited
growth of Palmer amaranth at second cohort is likely a reduced plant
response to day length, light availability and thermal units (e.g,
growing degree days). It is hypothesize that reduced day length
contributed to smaller plants at second cohort as well as shorter
flowering period. A study in North Carolina and Illinois predicted that
less than 10\% Palmer amaranth seedlings emergence occurred after June
(Piskackova et al., 2021). In addition, Palmer amaranth negative impact
on soybean (Korres et al., 2020) and cotton (Webster and Grey, 2015)
yields was higher when plants were established near to crop planting.
